PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-05-10 peace/papers/NAR_web_server_2009/paper.tex
Final version of supplementary materials
Modification of manuscript for final submission

\usepackage[usenames,dvipsnames]{color}
\definecolor{red}{rgb}{1,0,0}
\newcommand{\mc}[1]{\textcolor{red}{#1}}

@@

% Some shortcuts to make life easier when writing latex
\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}
\newcommand{\easycluster} {{\small EasyCluster}}
\newcommand{\metasim} {{\small MetaSim}}

@@

\begin{abstract}
  \mc{We present \peace, a stand alone tool for high-throughput {\it
      ab initio} clustering of transcript fragment sequences produced
    by Next Generation or Sanger Sequencing technologies.  It is
    freely available from
    \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
    and managed through a download-able user-friendly GUI, \peace\/
    can process large data sets of transcript fragments of length 50
    bases or greater, grouping the fragments by gene associations with
    a sensitivity compatible to leading clustering tools.  Once
    clustered, the user can employ the GUI's analysis functions,
    facilitating the easy collection of statistics and allowing them
    to single out specific clusters for more comprehensive study or
    assembly.  Using a novel {\it minimum spanning tree} based
    clustering method, \peace\/ is the equal of other leading tools in
    the literature in quality with an interface making it
    accessible to any user.  It produces results of quality virtually
    identical to those of the \wcd\/ tool when applied to Sanger
    sequences, significantly improved results over \wcd\/ when applied
    the products of Next Generation Sequencing Technology, and
    significantly improved results over \capthree\/ in both cases.  In
    short, \peace\/ provides an intuitive graphical user interface and
    a feature-rich, parallel clustering engine that proves to be a
    valuable addition to the leading cDNA clustering tools.}

@@

In the past, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
(NGS) technology has begun to rapidly replace Sanger Sequencing,
allowing for more comprehensive coverage of the transcriptome.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher rate of coverage than Sanger Sequencing
\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets -- and thus a
need for fast, efficient analysis algorithms.

@@

Given a set of transcript fragments sampled from across the genome, a
necessary first step of the set's analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time \cite{Hazelhurst08a}.

However, clustering is a computationally challenging problem.  Even
with the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster on the basis of pair-wise
sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS technologies exacerbates
this problem.  To deal with this, \peace\/ combines our own version of
the $d^2$ alignment-free sequence distance function \cite{Hide94} and
the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
and accurately find clusters of ESTs expressed from the same gene
without reference to a sequenced genome.  \mc{Compared against
  clustering tool in the literature
  \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
    Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
  sensitive and robust than the other tools, and more user friendly
  than any published clustering tool.  Nor are any of} \mc{these tools
  designed for the ease of installation and use that \peace\/
  provides.}

@@

The \peace\/ GUI can be launched by downloading and executing the
Jar file available on the \peace\/ website
(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
machine running the standard Java Virtual Machine (JVM).  Once
running, the user can employ the GUI to install the clustering engine
and perform a clustering of a data file in {\sc fasta} format, view an
initial analysis of the clusters, and produce files containing subsets
of the clusters as input to assembly tools such as \capthree\/ 
\cite{Huang99}.  A typical (first) use of \peace\/ must be
performed in the following manner (see Figure~\ref{fig:workflow}):

@@

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
  \cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).
  We notice essentially identical results for \peace\/ and wcd\/ in
  quality, both significantly better than \capthree\/ in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with \peace\/ showing a
  60\% runtime improvement over \wcd\/ in the first data set, but requiring 20\%
  more time than \wcd\/ in the Arabidopsis data set.  (We cannot make
  a reasonable runtime comparison to \capthree, as the runtime of that
  tool reflects both clustering and assembly.)  In Table~S1 of the
  Supplementary Materials we present runtimes of several more sets,
  observing that while \peace\/ appears to be significantly 
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}

@@

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, other natural biological effects (e.g. the trans-splicing of
transcripts), effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  These problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small cluster (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.

\mc{Peace can be downloaded from
\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
committed to keep maintaining and improving the tool in the future.
Meanwhile, we are developing our own MST-based assembly tool that can
seamlessly integrate with PEACE.  The underlying modular design of
PEACE offers users many possibilities to expand and incorporate the
MST algorithm for other bioinformatics application.}

>>>>

\usepackage[usenames,dvipsnames]{color}
\definecolor{red}{rgb}{1,0,0}
\newcommand{\mc}[1]{#1}

@@

% Some shortcuts to make life easier when writing latex
\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}
\newcommand{\easycluster} {{\small EasyCluster}}
\newcommand{\metasim} {{\small MetaSim}}
\newcommand{\tgicl} {{\small TGICL}}

@@

\begin{abstract}
  \mc{We present \peace, a stand alone tool for high-throughput {\it
      ab initio} clustering of transcript fragment sequences produced
    by Next Generation or Sanger Sequencing technologies.  It is
    freely available from
    \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
    and managed through a download-able user-friendly GUI, \peace\/
    can process large data sets of transcript fragments of length 50
    bases or greater, grouping the fragments by gene associations with
    a sensitivity compatible to leading clustering tools.  Once
    clustered, the user can employ the GUI's analysis functions,
    facilitating the easy collection of statistics and allowing them
    to single out specific clusters for more comprehensive study or
    assembly.  Using a novel {\it minimum spanning tree} based
    clustering method, \peace\/ is the equal of other leading tools in
    the literature in quality with an interface making it
    accessible to any user.  It produces results of quality virtually
    identical to those of the \wcd\/ tool when applied to Sanger
    sequences, significantly improved results over \wcd\/ and \tgicl\/ when applied
    the products of Next Generation Sequencing Technology, and
    significantly improved results over \capthree\/ in both cases.  In
    short, \peace\/ provides an intuitive graphical user interface and
    a feature-rich, parallel clustering engine that proves to be a
    valuable addition to the leading cDNA clustering tools.}

@@

In the past, access to a transcriptome sequence was primarily through
the use of Expressed Sequence Tags (ESTs), single-pass cDNA sequences
derived from transcribed mRNAs and sequenced by Sanger Sequencing
technology.  More recently, Next Generation Sequencing (NGS)
technology has begun to rapidly replace Sanger Sequencing.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher \cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets -- and thus a
need for fast, efficient analysis algorithms.

@@

Given a set of transcript fragments sampled from across the genome, a
necessary first step of the set's analysis is clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time \cite{Hazelhurst08a}.

Clustering is a computationally challenging problem; the runtime and
memory requirements to cluster on the basis of pair-wise sequence
alignments make such an approach infeasible in practice.  To deal with
this, \peace\/ combines our own version of the $d^2$ alignment-free
sequence distance function \cite{Hide94} and the concept of a {\it
  minimum spanning tree} \cite{Prim57} to quickly and accurately find
clusters of ESTs expressed from the same gene without reference to a
sequenced genome.  Compared against clustering tool in the literature
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
of equivalent quality to the \wcd\/ and \tgicl tools
\cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
other tools.  From the user perspective, o tool in the literature can
match \peace for ease of instillation, use, or the post-clustering
analysis tools \peace provides.

@@

The \peace\/ GUI can be launched by through the
Jar file available on the \peace\/ website
(\href{http://www.peace-tools.org}{www.peace-tools.org}) on any
machine running the standard Java Virtual Machine (JVM).  Once
running, the user can employ the GUI to install the clustering engine
and perform a clustering of a data file in {\sc fasta} format, view an
initial analysis of the clusters, and produce files containing subsets
of the clusters as input to assembly tools such as \capthree\/ 
\cite{Huang99}.  A typical (first) use of \peace\/ must be
performed in the following manner (see Figure~\ref{fig:workflow}):

@@

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and a 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

While \peace\/ clearly matches \wcd\/ on long-read sequence, comparisons
against \tgicl\/ \cite{Pertea2003}  are more complicated.  In short, we find that when
parameters are picked appropriately, \tgicl\/ is more sensitive at lower
error rate -- but \peace\/ is more robust to error, in all cases
has an improved Type 2 error rates and is better able to
distinguish duplicated genes.  (See the Supplementary Materials for a
more detailed analysis.)

In applying the tools to real data Sanger data, we used the Human
Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
\cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).  We
notice essentially identical results for \peace\/ and wcd\/ in
quality, both significantly better than \capthree\/ in Sensitivity and
Type 1 error rate, while slightly worse in the Jaccard Index and Type
2 error.  In runtime we see some inconsistency, with \peace\/ showing
a 60\% runtime improvement over \wcd\/ in the first data set, but
requiring 20\% more time than \wcd\/ in the Arabidopsis data set.  (We
cannot make a reasonable runtime comparison to \capthree, as the
runtime of that tool reflects both clustering and assembly.)  In
Table~S1 of the Supplementary Materials we present runtimes of several
more sets, observing that while \peace\/ appears to be significantly
faster on the smaller sets, \wcd\/ does overtake it for larger sets.

@@

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, other natural biological effects (e.g. the trans-splicing of
transcripts), effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  These problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small clusters (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.

Peace can be downloaded from
\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
committed to keep maintaining and improving the tool in the future.
Meanwhile, we are developing our own MST-based assembly tool that can
seamlessly integrate with PEACE.  The underlying modular design of
PEACE offers users many possibilities to expand and incorporate the
MST algorithm for other bioinformatics application.
